### 2019-01-21 KineticsPaperArXiv/assa.tex
..

Herein, we use a random vector 
$\vec\xi \in \Omega\in\mathbb{R}^{N_p}$ to parameterize model uncertainties, 
where $N_p$ is the number of uncertain inputs.
In practical computations, the \emph{canonical} variables $\xi_i, i=1,\ldots ,N_p$, are  
mapped to physical ranges meaningful in a given mathematical model. 
As mentioned in the introduction, an active subspace is a low-dimensional subspace
that consists of important directions in a model's input
parameter space~\cite{Constantine:2015}. The effective variability in a model output $f$
due to uncertain inputs is predominantly captured
along these directions. 
The directions constituting the active subspace are the dominant eigenvectors of the positive
semidefinite matrix 
%
\be
\mat{C} = \int_\Omega (\nabla_{\vec{\xi}}f)(\nabla_{\vec{\xi}}f)^\top \mu(d\vec\xi), 
\label{eq:C}
\ee
%
with 
$\mu(d\vec{\xi}) = \pi(\vec{\xi})d\vec{\xi}$, where $\pi(\vec{\xi})$ is the joint probability
\rebut{density function} of $\vec{\xi}$. Herein, $f$ is assumed to be a square integrable 
function with continuous partial 
derivatives with respect to the input parameters; moreover, we assume the partial derivatives
are square integrable. 
%Hence, it is possible that a given $f(\vec{\xi})$ might not admit an active
%subspace. However, it is of remarkable interest to investigate if one exists
%since the subspace could be exploited to reduce the dimensionality of the
%problem and hence the associated computational effort.
%Here,
%$\nabla_{\vec{\xi}}f$ denotes the gradient vector. 
%with individual components
%being partial derivatives of $f$ with respect to the $i^\text{th}$ input, $\xi_i$. 
Since $\mat{C}$ is symmetric and
positive semidefinite, it admits a spectral decomposition:
%
\be
\mat{C} = \mat{W}\mat{\Lambda}\mat{W}^\top.
\ee
%
Here $\mat{\Lambda}$ = diag($\lambda_1,\ldots,\lambda_{N_p}$) with the eigenvalues
$\lambda_i$'s sorted in descending order
\[
     \lambda_1 \geq \lambda_2 \geq \cdots \geq \lambda_\Np \geq 0,
\] 
and $\mat{W}$ has the (orthonormal) eigenvectors $\vec{w}_1, \ldots, \vec{w}_\Np$ as its columns.
The eigenpairs are partitioned about the $r$th eigenvalue such that
$\lambda_r/\lambda_{r+1}\gg 1$, 
\be
 \mat{W} = [\mat{W}_1~\mat{W}_2],~~\mat{\Lambda} = \begin{bmatrix}\mat{\Lambda}_1 & \\  &
  \mat{\Lambda}_2. 
\end{bmatrix}
\ee
 %
The columns of $\mat{W}_1 = 
\begin{bmatrix} \vec{w}_\rebut{1} \cdots \vec{w}_r\end{bmatrix}$ 
span the dominant eigenspace of $\mat{C}$ and
define the active subspace, and $\mat{\Lambda}_1$ is a diagonal matrix with
the corresponding set of eigenvalues, $\lambda_1, \ldots, \lambda_r$, on its diagonal. 
Once the active subspace
is computed, dimension reduction is accomplished by transforming the parameter
vector $\vec\xi$ into 
$\vec{y} = \mat{W}_1^\top\vec{\xi} \in \R^r$. The elements of $\vec{y}$ are 
referred to as the set of active variables. 
%The number $r$ of active
%variables is equal to the number of dominant eigenvectors.

@@

\section{GSA measures and their links with active subspaces}
\label{sub:gsa}
Consider a function $f = f(\xi_1, \xi_2, \ldots, \xi_\Np)$. 
While the active subspace framework described above does not make any assumptions
about independence of the inputs $\xi_i$, $i = 1, \ldots, \Np$, 
the classical 
framework of variance based sensitivity analysis~\cite{Sobol:2001, Saltelli:2010} 
assumes that the inputs
are statistically independent. While extensions to the cases 
of correlated inputs exist~\cite{Borgonovo:2007,Li:2010,Jacques:2006,Xu:2007},
 we limit the discussion in this section to the
case of random inputs that are statistically independent and are 
either uniformly distributed or
distributed according to the Boltzmann probability distribution.
Note that a measure $\mu$
on $\R$ is referred to as a Boltzmann measure if it is 
absolutely continuous with respect to the Lebesgue measure  
and admits a density  of the form $\pi(x) = C \exp\{-V(x)\}$,
where $V$ is a continuous function and $C$ a normalization 
constant~\cite{Lamboni:2013}. \rebut{
An important class of Boltzmann distributions are the so called log-concave
distributions, which include Normal, Exponential, Beta, Gamma, Gumbel, and
Weibull distributions. Note also that the uniform distribution does not fall under
the class of Boltzmanm distributions~\cite{Lamboni:2013}.}

The total-effect Sobol' index ($T_i(f)$) of a model output, $f(\vec\xi)$ quantifies
the total contribution of the input, $\xi_i$ to the variance of the
output~\cite{Sobol:2001}. Mathematically, this can be expressed as follows:
%
\be
T_i(f) = 1 - 
\frac{\V_{\rebut{\vec{\xi}_{\sim i}}}\big[\mathbb{E}[f|\vec{\xi}_{\sim i}]\big]}{\V(f)},
\label{eq:total}
\ee
%
where $\vec{\xi}_{\sim i}$ is the input parameter vector with the  
$i^\text{th}$ entry removed. \rebut{
Here $\mathbb{E}[f|\vec{\xi}_{\sim i}]$ denotes the conditional
expectation of $f$ given $\vec{\xi}_{\sim i}$
and its variance is computed with respect to $\vec{\xi}_{\sim i}$.}
The quantity, $\V(f)$ denotes the total variance of the model output.
 The total-effect Sobol' index accounts
for the contribution of a given input to the variability in the output by itself
as well as due to its interaction or coupling with other inputs. 
Determining accurate estimates of $T_i(f)$ typically involves a large 
number of 
model runs and is therefore can be prohibitive in the case of
compute-intensive applications. Derivative based 
global sensitivity measures (DGSMs)~\cite{Sobol:2009} provide a means for
approximating informative upper bounds on $T_i(f)$ at a lower cost; see 
also~\cite{Vohra:2018}.

For $f: \Omega \to \R$, we consider the DGSMs,
\[
    \nu_i(f) := \E{\left(\frac{\partial f}{\partial\xi_i}\right)^2} =
                  \int_\Omega 
                  \left(\frac{\partial f}{\partial\xi_i}\right)^2
                  \pi(\vec{\xi})d\vec{\xi}, \quad i = 1, \ldots, \Np.   
\]
Here $\pi$ is the joint PDF of $\vec\xi$. \rebut{Note that $\nu_i(f)$ is the 
$i^{\text{th}}$ diagonal element of the matrix $\mat{C}$} as defined in~\eqref{eq:C}. 
Consider the spectral decomposition written 
as $\mat{C} = \sum_{k=1}^\Np \lambda_k \vec{w}_k \vec{w}_k^\top$. Herein, we use the notation 
$\ip{\cdot}{\cdot}$ for the Euclidean inner product.
% where
%$\lambda_i$ are the (non-negative) eigenvalues of $\mat{C}$, in descending
%order, and
%$\vec{w}_k$ are the corresponding (orthonormal) eigenvectors. 
The following
result provides a representation of DGSMs in terms of the 
spectral representation of $\mat{C}$: 
\begin{lemma}
We have
$\nu_i(f) = \sum_{k=1}^\Np \lambda_k \ip{\vec{e}_i}{\vec{w}_k}^2$.
\end{lemma}
\begin{proof}
Note that $\nu_i(f) = \vec{e}_i^\top \mat{C} \vec{e}_i$,  
where $\vec{e}_i$ is the $i$th coordinate vector in $\R^\Np$, $i = 1, \ldots, \Np$.
Therefore,
$\nu_i(f) = \vec{e}_i^T \Big(\sum_{k=1}^\Np \lambda_k \vec{w}_k \vec{w}_k^\top\Big) \vec{e}_i
 = \sum_{k=1}^\Np \lambda_k \ip{\vec{e}_i}{\vec{w}_k}^2$. 
\end{proof}
In the case where the eigenvalues decay rapidly to zero, we can obtain
accurate approximations of $\nu_i(f)$ by truncating the summation: 
\[
   \act{f}{i}{r} =  \sum_{k=1}^r \lambda_k \ip{\vec{e}_i}{\vec{w}_k}^2,
   \quad i = 1, \ldots, \Np, \quad r \leq \Np.
\]
The quantities $\act{f}{i}{r}$ are called activity scores
in~\cite{Diaz:2016,Constantine:2017}, where links between GSA measures and
active subspaces is explored.
%The activity scores connect ideas from active subspaces and global sensitivity
%analysis, and can be used to approximate DGSMs.  
The following result, which
can also be found in~\cite{Diaz:2016,Constantine:2017}, quantifies the error in this
approximation. We provide a short proof for completeness. 
\begin{proposition}\label{prp:dgsm_bound} 
For $1 \leq r < \Np$,
\[
0 \leq \nu_i(f) - \act{f}{i}{r} \leq \lambda_{r+1}, \quad i = 1, \ldots, \Np.
\] 
\end{proposition}
\begin{proof}

>>>>

Herein, we use a random vector 
$\vec\xi \in \Omega\in\mathbb{R}^{N_p}$ to parameterize model uncertainties, 
where $N_p$ is the number of uncertain inputs.
In practical computations, the \emph{canonical} variables $\xi_i, i=1,\ldots ,N_p$, are  
mapped to physical ranges meaningful in a given mathematical model. 
As mentioned in the introduction, an active subspace is a low-dimensional subspace
that consists of important directions in a model's input
parameter space~\cite{Constantine:2015}. The effective variability in a model output $f$
due to uncertain inputs is predominantly captured
along these directions. 
The directions constituting the active subspace are the dominant eigenvectors of the positive
semidefinite matrix 
%
\be
\mat{C} = \int_\Omega (\nabla_{\vec{\xi}}f)(\nabla_{\vec{\xi}}f)^\top \mu(d\vec\xi), 
\label{eq:C}
\ee
%
with 
$\mu(d\vec{\xi}) = \pi(\vec{\xi})d\vec{\xi}$, where $\pi(\vec{\xi})$ is the joint probability
density function of $\vec{\xi}$. Herein, $f$ is assumed to be a square integrable 
function with continuous partial 
derivatives with respect to the input parameters; moreover, we assume the partial derivatives
are square integrable. 
%Hence, it is possible that a given $f(\vec{\xi})$ might not admit an active
%subspace. However, it is of remarkable interest to investigate if one exists
%since the subspace could be exploited to reduce the dimensionality of the
%problem and hence the associated computational effort.
%Here,
%$\nabla_{\vec{\xi}}f$ denotes the gradient vector. 
%with individual components
%being partial derivatives of $f$ with respect to the $i^\text{th}$ input, $\xi_i$. 
Since $\mat{C}$ is symmetric and
positive semidefinite, it admits a spectral decomposition:
%
\be
\mat{C} = \mat{W}\mat{\Lambda}\mat{W}^\top.
\ee
%
Here $\mat{\Lambda}$ = diag($\lambda_1,\ldots,\lambda_{N_p}$) with the eigenvalues
$\lambda_i$'s sorted in descending order
\[
     \lambda_1 \geq \lambda_2 \geq \cdots \geq \lambda_\Np \geq 0,
\] 
and $\mat{W}$ has the (orthonormal) eigenvectors $\vec{w}_1, \ldots, \vec{w}_\Np$ as its columns.
The eigenpairs are partitioned about the $r$th eigenvalue such that
$\lambda_r/\lambda_{r+1}\gg 1$, 
\be
 \mat{W} = [\mat{W}_1~\mat{W}_2],~~\mat{\Lambda} = \begin{bmatrix}\mat{\Lambda}_1 & \\  &
  \mat{\Lambda}_2. 
\end{bmatrix}
\ee
 %
The columns of $\mat{W}_1 = 
\begin{bmatrix} \vec{w}_1 \cdots \vec{w}_r\end{bmatrix}$ 
span the dominant eigenspace of $\mat{C}$ and
define the active subspace, and $\mat{\Lambda}_1$ is a diagonal matrix with
the corresponding set of eigenvalues, $\lambda_1, \ldots, \lambda_r$, on its diagonal. 
Once the active subspace
is computed, dimension reduction is accomplished by transforming the parameter
vector $\vec\xi$ into 
$\vec{y} = \mat{W}_1^\top\vec{\xi} \in \R^r$. The elements of $\vec{y}$ are 
referred to as the set of active variables. 
%The number $r$ of active
%variables is equal to the number of dominant eigenvectors.

@@

\section{GSA measures and their links with active subspaces}
\label{sub:gsa}
Consider a function $f = f(\xi_1, \xi_2, \ldots, \xi_\Np)$. 
While the active subspace framework described above does not make any assumptions
about independence of the inputs $\xi_i$, $i = 1, \ldots, \Np$, 
the classical 
framework of variance based sensitivity analysis~\cite{Sobol:2001, Saltelli:2010} 
assumes that the inputs
are statistically independent. While extensions to the cases 
of correlated inputs exist~\cite{Borgonovo:2007,Li:2010,Jacques:2006,Xu:2007},
 we limit the discussion in this section to the
case of random inputs that are statistically independent and are 
either uniformly distributed or
distributed according to the Boltzmann probability distribution.
Note that a measure $\mu$
on $\R$ is referred to as a Boltzmann measure if it is 
absolutely continuous with respect to the Lebesgue measure  
and admits a density  of the form $\pi(x) = C \exp\{-V(x)\}$,
where $V$ is a continuous function and $C$ a normalization 
constant~\cite{Lamboni:2013}. 
An important class of Boltzmann distributions are the so called log-concave
distributions, which include Normal, Exponential, Beta, Gamma, Gumbel, and
Weibull distributions. Note also that the uniform distribution does not fall under
the class of Boltzmanm distributions~\cite{Lamboni:2013}.

The total-effect Sobol' index ($T_i(f)$) of a model output, $f(\vec\xi)$ quantifies
the total contribution of the input, $\xi_i$ to the variance of the
output~\cite{Sobol:2001}. Mathematically, this can be expressed as follows:
%
\be
T_i(f) = 1 - 
\frac{\V_{\vec{\xi}_{\sim i}}\big[\mathbb{E}[f|\vec{\xi}_{\sim i}]\big]}{\V(f)},
\label{eq:total}
\ee
%
where $\vec{\xi}_{\sim i}$ is the input parameter vector with the  
$i^\text{th}$ entry removed. 
Here $\mathbb{E}[f|\vec{\xi}_{\sim i}]$ denotes the conditional
expectation of $f$ given $\vec{\xi}_{\sim i}$
and its variance is computed with respect to $\vec{\xi}_{\sim i}$.
The quantity, $\V(f)$ denotes the total variance of the model output.
 The total-effect Sobol' index accounts
for the contribution of a given input to the variability in the output by itself
as well as due to its interaction or coupling with other inputs. 
Determining accurate estimates of $T_i(f)$ typically involves a large 
number of 
model runs and is therefore can be prohibitive in the case of
compute-intensive applications. Derivative based 
global sensitivity measures (DGSMs)~\cite{Sobol:2009} provide a means for
approximating informative upper bounds on $T_i(f)$ at a lower cost; see 
also~\cite{Vohra:2018}.

For $f: \Omega \to \R$, we consider the DGSMs,
\[
    \nu_i(f) := \E{\left(\frac{\partial f}{\partial\xi_i}\right)^2} =
                  \int_\Omega 
                  \left(\frac{\partial f}{\partial\xi_i}\right)^2
                  \pi(\vec{\xi})d\vec{\xi}, \quad i = 1, \ldots, \Np.   
\]
Here $\pi$ is the joint PDF of $\vec\xi$. Note that $\nu_i(f)$ is the 
$i^{\text{th}}$ diagonal element of the matrix $\mat{C}$ as defined in~\eqref{eq:C}. 
Consider the spectral decomposition written 
as $\mat{C} = \sum_{k=1}^\Np \lambda_k \vec{w}_k \vec{w}_k^\top$. Herein, we use the notation 
$\ip{\cdot}{\cdot}$ for the Euclidean inner product.
% where
%$\lambda_i$ are the (non-negative) eigenvalues of $\mat{C}$, in descending
%order, and
%$\vec{w}_k$ are the corresponding (orthonormal) eigenvectors. 
The following
result provides a representation of DGSMs in terms of the 
spectral representation of $\mat{C}$: 
\begin{lemma}
We have
$\nu_i(f) = \sum_{k=1}^\Np \lambda_k \ip{\vec{e}_i}{\vec{w}_k}^2$.
\end{lemma}
\begin{proof}
Note that $\nu_i(f) = \vec{e}_i^\top \mat{C} \vec{e}_i$,  
where $\vec{e}_i$ is the $i$th coordinate vector in $\R^\Np$, $i = 1, \ldots, \Np$.
Therefore,
$\nu_i(f) = \vec{e}_i^T \Big(\sum_{k=1}^\Np \lambda_k \vec{w}_k \vec{w}_k^\top\Big) \vec{e}_i
 = \sum_{k=1}^\Np \lambda_k \ip{\vec{e}_i}{\vec{w}_k}^2$. 
\end{proof}
In the case where the eigenvalues decay rapidly to zero, we can obtain
accurate approximations of $\nu_i(f)$ by truncating the summation: 
\[
   \act{f}{i}{r} =  \sum_{k=1}^r \lambda_k \ip{\vec{e}_i}{\vec{w}_k}^2,
   \quad i = 1, \ldots, \Np, \quad r \leq \Np.
\]
The quantities $\act{f}{i}{r}$ are called activity scores
in~\cite{Diaz:2016,Constantine:2017}, where links between GSA measures and
active subspaces is explored.
%The activity scores connect ideas from active subspaces and global sensitivity
%analysis, and can be used to approximate DGSMs.  
The following result, which
can also be found in~\cite{Diaz:2016,Constantine:2017}, quantifies the error in this
approximation. We provide a short proof for completeness. 
\begin{proposition}\label{prp:dgsm_bound} 
For $1 \leq r < \Np$,
\[
0 \leq \nu_i(f) - \act{f}{i}{r} \leq \lambda_{r+1}, \quad i = 1, \ldots, \Np.
\] 
\end{proposition}
\begin{proof}
